taxadb: A High-Performance Local Taxonomic Database Interface

### 2020-05-25 paper/manuscript.Rmd
reviewer responses, additional citations

---
title: "taxadb: A High-Performance Local Taxonomic Database Interface"
author:
  - name: "Kari E A Norman"
    affiliation: ucb
  - name: "Scott Chamberlain"
    affiliation: ropensci
  - name: "Carl Boettiger"
    affiliation: ucb, ropensci
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
abstract: |
  A familiar and growing challenge in ecological and evolutionary research is that of reconciling scientific names of relevant taxa when combining data from separate sources. While this problem is already well understood and numerous naming authorities have been created to address the issue, most researchers lack a fast, consistent, and intuitive way to reconcile taxonomic names.  Here, we present `taxadb` R package to address this gap.  In contrast to existing tools, `taxadb` provides the following:  
  1) `taxadb` accesses established naming authorities to resolve synonyms, IDs, and hierarchical classification.  
  2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.  
  3) `taxadb` provides a consistent and intuitive data format.  
  4) `taxadb` is built on a simple, extensible and language agnostic design that can easily accommodate new authorities.

@@

While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008; @maldonado2015],
and a growing number of databases and tools have emerged over the past few decades
[e.g. @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018],
it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
Here, we present an R package, `taxadb`, which seeks to address this gap.

Databases of taxonomic names such as the Integrated Taxonomic Information System [ITIS; @itis], 
the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi], 
the Catalogue of Life [COL; @col], and over one hundred other providers have sought to address these problems
by providing expert-curated lists of accepted taxonomic names, synonyms, associated taxonomic rank,
hierarchical classification, and scientific authority (e.g. author and date) establishing a scientific name.
The R language [@R] is widely used in ecology and evolution [@Lai2019] and the `taxize` package [@Chamberlain2013]
has become a popular way for R users to interact with naming providers and name resolution services. 
`taxize` implements bindings to the web APIs (Application Programming Interface)
hosted by many popular taxonomic name providers. 
Unfortunately, this means that functions in the `taxize` are impacted by several major drawbacks
that are inherent in the implementation of these central API servers, such as:

- Queries require internet access at all times.
- Queries are slow and inefficient to implement and perform; frequently requiring separate API calls for each taxonomic name.
- The type of query is highly limited by the API design. For instance, it is usually impossible to make queries across the entire corpus of names, such as "which accepted name has the most known synonyms?".
- Both query formats and responses differ substantially across different naming providers, making it difficult to apply a script designed for one provider to different provider.
- Most queries are not reproducible, as the results depend on the state of the central server (and potentially the quality of the internet connection).  Many names providers update the server data either continuously or at regular intervals, including both revising existing names and adding new names.

Instead of binding existing web APIs, `taxadb` is built around a set of compressed text files which are 
automatically downloaded, imported, and stored on a local database by `taxadb`. 
The largest of the taxonomic naming providers today contain under 6 million name records with uncompressed file sizes
over a GB, which can be compressed to around 50 MB and downloaded in under a minute on a 1 MB/s connection. 
By using a local database as the backend, `taxadb` allows R users to interact with large data files without large memory (RAM) requirements.  A query for a single name over the web API requires a remote server to respond, execute the query, 
and serialize the response, which can take several seconds. Thus it does not take many taxa before transferring the 
entire data set to query locally is more efficient.  Moreover, this local copy can be cached on the user's machine, 
requiring only the one-time setup, and enabling offline use and reproducible queries.  Rather than returning data
in whatever format is given by the provider, `taxadb` provides a data structure following a consistent,
standardized layout or schema following the Darwin Core Standard [@Wieczorek2012].  Table 1 summarizes the list of all naming providers currently accessed by `taxadb`.  More details are provided in the Data Sources Vignette, <https://docs.ropensci.org/taxadb/articles/data-sources.html>.

@@

After loading the package, we look up the taxonomic identifier for Atlantic Cod, *Gadus morhua*, and the compliment:

@@

Taxonomic identifiers provide a fundamental abstraction which lies at the heart of managing taxonomic names.  For instance, by resolving scientific names to identifiers, we can identify which names are synonyms -- different scientific names used to describe the same species -- and which names are not recognized. Unmatched names may indicate an error in data entry or otherwise warrant further investigation. Taxon identifiers are also easily resolved to the original authority (scientific publication) establishing the name. (The historical practice of appending an author and year to a scientific name, e.g. *Poa annua ssp. annua* (Smith 1912), serves a valuable role in disambiguating different uses of the same name but can be notoriously harder to resolve to the appropriate reference, while variation in this convention creates many distinct versions of the same name [@Patterson2010]).

@@

```{r iucn_table, echo=FALSE, cache = FALSE}
status_data %>% printtable(caption = "The subset of the IUCN status data used for subsequent taxonomic identifier examples.") 
```

```{r trait_table, echo = FALSE, cache = FALSE }
trait_data %>% printtable(caption = "The subset of the Elton trait data used for subsequent taxonomic identifier examples.")
```

If we attempted to join these data directly on the species names provided by each table, we would find very little overlap, with only one species having both a body mass and an IUCN threat status resolved (Table 4).

@@

```{r echo = FALSE, cache = FALSE}
joined %>%
  printtable(caption = "Example IUCN and trait data joined directly on scientific name showing only one match. While common, joining on scientific name does not account for taxonomic inconsistencies between databases and therefore results in seemingly very little overlap in species representation between the two.")
```

@@

```{r cache = FALSE, echo = FALSE}
## Just for pretty-printing
joined %>%  
  tidyr::replace_na(list(category = "-", elton_name = "-", iucn_name = "-")) %>%
  select(elton_name, iucn_name, mass, category, id) %>%
  printtable(caption = "Example IUCN and trait data joined on taxonomic ID. Multiple species have a different scientific name in the Elton and IUCN Redlist databases but can be match based on their COL taxonomic ID.")
```

@@

`get_ids()` returns the `acceptedNameUsageID`, the identifier associated with the *accepted* name.  Some naming providers, such as ITIS and NCBI, provide taxonomic identifiers to both synonyms and accepted names.  Other providers, such as COL and GBIF, only provide identifiers for accepted names.  Common practice in Darwin Core archives is to provide an `acceptedNameUsageID` only for names which are synonyms, and otherwise to provide a `taxonID`.  For accepted names, the `acceptedNameUsageID` is then given as missing (`NA`), while for synonyms, the `taxonID` may be missing (`NA`).  In contrast, `taxadb` lists the `acceptedNameUsageID` for accepted names (where it matches the `taxonID`), as well as known synonyms.  This is semantically identical, but also more convenient for database interfaces, since it allows a name to mapped to it's accepted identifier (or an identifier to map to it's accepted name usage) without the additional logic.  For consistency, we will use the term "identifier" to mean the `acceptedNameUsageID` rather than the more ambiguous `taxonID` (which is undefined for synonyms listed by many providers), unless explicitly stated otherwise.

@@

`get_ids` offers a first pass at matching scientific names to id, but names may remain unresolved for a number of reasons. First, a name may match to multiple accepted names, as in the case of a species that has been split. By design, these cases are left to be resolved by the researcher using the `filter_` functions to filter underlying taxonomic tables for additional information. A name may also be unresolved due to typos or improper formatting. `clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g. `Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the binomial name. These modifications are not appropriate in all settings and should be used with care. Spell check of input names is outside the scope of `taxadb`.

Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym. Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@franz2009; @Franz2018]. The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".

@@

```{r}
filter_rank("Aves", rank="class", provider = "col") %>%
  pull(taxonID) %>%
  n_distinct()
```

@@

```{r echo = FALSE, cache = FALSE}
multi_match %>%
  select(1:5) %>%
  printtable(caption = "Some names may not resolve to an identifier using get\\_ids() because they match to more than one accepted ID. In such cases filter\\_ functions give further detail, as in the example of Zosterops rendovae below which has two accepted ID matches.", escape = FALSE)
```

@@

`taxadb` can also resolve common names to their identifier by mapping common name to the accepted scientific name. Common names come with many of the same problems as scientific names but often more frequent, e.g. matching to more than one accepted name, non-standardized formatting (like capitalization) and spelling etc. Common names are accessed via `filter_common` which takes a vector of common names. The user can then resolve discrepancies.

@@

Taxonomic identifiers are a powerful tool for maintaining taxonomic consistency, an essential task for a wide variety of applications. Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary for resolving even moderately sized data sets. `taxadb` fills an important gap between existing tools and typical research patterns by providing a fast, reproducible approach for matching names to taxonomic identifiers. `taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution. In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.

>>>>

---
title: "taxadb: A High-Performance Local Taxonomic Database Interface"
author:
  - name: "Kari E A Norman"
    affiliation: ucb
  - name: "Scott Chamberlain"
    affiliation: ropensci
  - name: "Carl Boettiger"
    affiliation: ucb, ropensci
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
abstract: |
  A familiar and growing challenge in ecological and evolutionary research is that of reconciling scientific names of relevant taxa when combining data from separate sources. While this problem is already well understood and numerous naming authorities have been created to address the issue, most researchers lack a fast, consistent, and intuitive way to reconcile taxonomic names.  Here, we present `taxadb` R package to address this gap.  In contrast to existing tools, `taxadb` provides all of the following in one tool:  
  1) `taxadb` accesses established naming authorities to resolve synonyms, taxonomic identifiers, and hierarchical classification.  
  2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.  
  3) `taxadb` provides a consistent and intuitive data format.  
  4) `taxadb` is built on a simple, extensible, and language-agnostic design that can easily accommodate new authorities.

@@

While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008; @maldonado2015; @remsen2016],
and a growing number of databases and tools have emerged over the past few decades
[e.g. @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018; @gries2014],
it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
Here, we present an R package, `taxadb`, which seeks to address this gap.

Databases of taxonomic names such as the Integrated Taxonomic Information System [ITIS; @itis], 
the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi], 
the Catalogue of Life [COL; @col], and over one hundred other providers have sought to address these problems
by providing expert-curated lists of accepted taxonomic names, synonyms, associated taxonomic rank,
hierarchical classifications, and scientific authority (e.g. author and date) establishing a scientific name.
The R language [@R] is widely used in ecology and evolution [@Lai2019] and the `taxize` package [@Chamberlain2013]
has become a popular way for R users to interact with naming providers and name resolution services. 
`taxize` implements bindings to the web APIs (Application Programming Interface)
hosted by many popular taxonomic name providers. 
Nevertheless, this means that functions in the `taxize` are impacted by several major drawbacks
that are inherent in the implementation of these central API servers, such as:

- Queries require internet access at all times.
- Queries are slow and inefficient to implement and perform; frequently requiring separate API calls for each taxonomic name.
- The type of query is highly limited by the API design. For instance, it is usually impossible to make queries across the entire corpus of names, such as "which accepted name has the most known synonyms?".
- Both query formats and responses differ substantially across different naming providers, making it difficult to apply a script designed for one provider to different provider.
- Most queries are not reproducible, as the results depend on the state of the central server (and potentially the quality of the internet connection)[@rees2017].  Many names providers update the server data either continuously or at regular intervals, including both revising existing names (for spelling or changes in accepted name designation) and adding new names.

Instead of binding existing web APIs, `taxadb` is built around a set of compressed text files which are 
automatically downloaded, imported, and stored on a local database by `taxadb`. 
The largest of the taxonomic naming providers today contain under 6 million name records with uncompressed file sizes
over a GB, which can be compressed to around 50 MB and downloaded in under a minute on a 1 MB/s connection. 
By using a local database as the backend, `taxadb` allows R users to interact with large data files without large memory (RAM) requirements.  A query for a single name over the web API requires a remote server to respond, execute the query, 
and serialize the response, which can take several seconds. Thus it does not take many taxa before transferring the 
entire data set to query locally is more efficient.  Moreover, this local copy can be cached on the user's machine, 
requiring only the one-time setup, and enabling offline use and reproducible queries.  Rather than returning data
in whatever format is given by the provider, `taxadb` provides a data structure following a consistent,
standardized layout or schema following Darwin Core, which provides standard terms for biodiversity data [@Wieczorek2012].  Table 1 summarizes the list of all naming providers currently accessed by `taxadb`.  More details are provided in the Data Sources Vignette, <https://docs.ropensci.org/taxadb/articles/data-sources.html>.

@@

After loading our package and the tidyverse package for ease in manipulating function output, we look up the taxonomic identifier for Atlantic Cod, *Gadus morhua*, and the compliment:

@@

Taxonomic identifiers provide a fundamental abstraction which lies at the heart of managing taxonomic names.  For instance, by resolving scientific names to identifiers, we can identify which names are synonyms -- different scientific names used to describe the same species -- and which names are not recognized. Unmatched names may indicate an error in data entry or otherwise warrant further investigation. Taxon identifiers are also easily resolved to the original authority (scientific publication) establishing the name. The common practice of appending an author and year to a scientific name, e.g. *Poa annua annua* (Smith 1912), serves a valuable role in disambiguating different uses of the same name but can be notoriously harder to resolve to the appropriate reference, while variation in this convention creates many distinct versions of the same name [@Patterson2010].

@@

```{r iucn_table, echo=FALSE, cache = FALSE}
status_data %>% printtable(caption = "The subset of the IUCN status data used for subsequent taxonomic identifier examples.") %>%
  column_spec(1, italic = TRUE)
```

```{r trait_table, echo = FALSE, cache = FALSE }
trait_data %>% printtable(caption = "The subset of the Elton trait data used for subsequent taxonomic identifier examples.") %>%
  column_spec(1, italic = TRUE)
```

If we attempted to join these data directly on the species names provided by each table, we would find very little overlap, with only one species name having both a body mass and an IUCN threat status resolved (Table 4).

@@

```{r echo = FALSE, cache = FALSE}
joined %>%
  printtable(caption = "Example IUCN and trait data joined directly on scientific name showing only one match. While common, joining on scientific name does not account for nomenclatural and taxonomic inconsistencies between databases and therefore results in seemingly very little overlap in species representation between the two.") %>%
  column_spec(1, italic = TRUE)
```

@@

```{r cache = FALSE, echo = FALSE}
## Just for pretty-printing
joined %>%  
  tidyr::replace_na(list(category = "-", elton_name = "-", iucn_name = "-")) %>%
  select(elton_name, iucn_name, mass, category, id) %>%
  printtable(caption = "Example IUCN and trait data joined on taxonomic ID. Multiple species have a different scientific name in the Elton and IUCN Redlist databases but can be match based on their COL taxonomic ID.") %>%
  column_spec(c(1,2), italic = TRUE)
```

@@

`get_ids()` returns the `acceptedNameUsageID`, the identifier associated with the *accepted* name.  Some naming providers, such as ITIS and NCBI, provide taxonomic identifiers to both synonyms and accepted names.  Other providers, such as COL and GBIF, only provide identifiers for accepted names.  Common practice in Darwin Core archives is to provide an `acceptedNameUsageID` only for names which are synonyms, and otherwise to provide a `taxonID`.  For accepted names, the `acceptedNameUsageID` is then given as missing (`NA`), while for synonyms, the `taxonID` may be missing (`NA`).  In contrast, `taxadb` lists the `acceptedNameUsageID` for accepted names (where it matches the `taxonID`), as well as known synonyms.  This is semantically identical, but also more convenient for database interfaces, since it allows a name to mapped to its accepted identifier (or an identifier to map to it's accepted name usage) without the additional logic.  For consistency, we will use the term "identifier" to mean the `acceptedNameUsageID` rather than the more ambiguous `taxonID` (which is undefined for synonyms listed by many providers), unless explicitly stated otherwise.

@@

`get_ids` offers a first pass at matching scientific names to id, but names may remain unresolved for a number of reasons. First, a name may match to multiple accepted names, as in the case of a species that has been split. By design, these cases are left to be resolved by the researcher using the `filter_` functions to filter underlying taxonomic tables for additional information. A name may also be unresolved due to typos or improper formatting. `clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g. `Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the binomial name. These modifications are not appropriate in all settings and should be used with care. Spell check of input names is outside the scope of `taxadb`, however existing tools such as those developed by the Global Names Architecture (http://globalnames.org/apps/) could be incorporated into a `taxadb` workflow.

Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym. Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@franz2009; @Franz2018; @lepage2014]. The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".

@@

```{r}
filter_rank("Aves", rank="class", provider = "col") %>%
  filter(taxonomicStatus == "accepted") %>%
  pull(taxonID) %>%
  n_distinct()
```

@@

```{r echo = FALSE, cache = FALSE}
multi_match %>%
  select(1:5) %>%
  printtable(caption = "Some names may not resolve to an identifier using get\\_ids() because they match to more than one accepted ID. In such cases filter\\_ functions give further detail, as in the example of *Zosterops rendovae* below which has two accepted ID matches.", escape = FALSE) %>%
  column_spec(3, italic = TRUE)
```

@@

`taxadb` can also resolve common names to their identifier by mapping common name to the accepted scientific name. Common names have many of the same issues as scientific names even more frequently (e.g. matching to more than one accepted name, non-standardized formatting). Common names are accessed via `filter_common` which takes a vector of common names. The user can then resolve discrepancies.

@@

Taxonomic identifiers are an essential first step for maintaining taxonomic consistency, a key task for a wide variety of applications. Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary for resolving even moderately sized data sets. `taxadb` fills an important gap between existing tools and typical research patterns by providing a fast, reproducible approach for matching names to taxonomic identifiers. `taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution. In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.
